The Blessing of FedAvg's Linear Speedup in Federated Learning

### 2020-05-31 source/sec_Nesterov.tex
updated nesterov section and proofs

We first show that the Nesterov accelerated FedAvg has $O(1/NT)$
convergence rate for $\mu$-strongly convex and $L$-smooth objectives.
The Nesterov Accelerated FedAvg algorithm follows the updates
\begin{align*}
y_{t+1}^{k} & =w_{t}^{k}-\alpha_{t}g_{t,k}\\
w_{t+1}^{k} & =\begin{cases}
y_{t+1}^{k}+\beta_{t}(y_{t+1}^{k}-y_{t}^{k}) & \text{if }t+1\notin\mathcal{I}_{E}\\
\sum_{k=1}^{N}p_{k}\left[y_{t+1}^{k}+\beta_{t}(y_{t+1}^{k}-y_{t}^{k})\right] & \text{if }t+1\in\mathcal{I}_{E}
\end{cases}
\end{align*}
where $g_{t,k}:=\nabla F_{k}(w_{t}^{k},\xi_{t}^{k})$ is the stochastic
gradient. 
\begin{theorem}
	Assume that $F_{k}$ is $L$-smooth and $\mu$-strongly convex for
	all $k$, and let $\nu_{\max}=\max_{k}\frac{1}{N}p_{k}$. Let $\kappa=\frac{L}{\mu}$,
	$\gamma=\max\{32\kappa,E\}$ where $E$ is the communication delay,
	and diminishing learning rates $\alpha_{t}=\beta_{t}=\frac{1}{4\mu(\gamma+t)}$.
	Let $\overline{w}_{T}=\sum_{k=1}^{N}p_{k}w_{T}^{k}$ be the average
	of local parameters at an arbitrary time $T$.

@@

In this section, we provide linear speedup analyses of Nesterov Accelerated
FedAvg with convex and smooth objectives and show that the optimality
gap is $O(1/\sqrt{NT})$ where $N$ is the number of participating
devices. This result complements the strongly convex case in the previous
part, as well as the non-convex smooth setting in {[}CITE kdd and
others{]}, where a similar $O(1/\sqrt{NT})$ rate is given in terms
of averaged gradient norm. 
\begin{theorem}
	Suppose $F_{k}$ is $L$-smooth and convex for all $k$, and let $\nu_{\max}=\max_{k}\frac{1}{N}p_{k}$.
	Set learning rate $\alpha_{t}=O(\sqrt{\frac{N}{T}})$.

	With full device participation, 
	\begin{align*}
	\min_{t\leq T}F(\overline{w}_{t})-F(w^{\ast}) & =O(\sqrt{\frac{\nu_{\max}^{2}\sigma^{2}}{NT}+\frac{E^{2}LG^{2}}{T^{4/3}}})
	\end{align*}
	and with partial device participation with $K$ sampled devices at
	each communication round, 
	\begin{align*}
	\min_{t\leq T}F(\overline{w}_{t})-F(w^{\ast}) & =O(\sqrt{\frac{\nu_{\max}^{2}\sigma^{2}}{NT}+\frac{E^{2}G^{2}}{KT}+\frac{E^{2}LG^{2}}{T^{4/3}}})
	\end{align*}
\end{theorem}
%
\begin{remark}
	It is possible to extend the analysis to accelerated FedAvg algorithms
	with other momentum-based updates. However, in line with the stochastic
	optimization setting, none of these methods will be able to achieve
	a better rate than the original FedAvg algorithm in general. For this
	reason, we will instead turn to the overparameterized setting {[}CITE
	Belkin{]} where we show that it is possible to improve the convergence
	rate of FedAvg with momentum based local updates. 
\end{remark}

>>>>

We first show that the Nesterov accelerated FedAvg has $O(1/NT)$
convergence rate for $\mu$-strongly convex and $L$-smooth objectives.
The Nesterov Accelerated FedAvg algorithm follows the updates
\begin{align*}
y_{t+1}^{k} & =w_{t}^{k}-\alpha_{t}g_{t,k}\\
w_{t+1}^{k} & =\begin{cases}
y_{t+1}^{k}+\beta_{t}(y_{t+1}^{k}-y_{t}^{k}) & \text{if }t+1\notin\mathcal{I}_{E}\\
\sum_{k=1}^{N}p_{k}\left[y_{t+1}^{k}+\beta_{t}(y_{t+1}^{k}-y_{t}^{k})\right] & \text{if }t+1\in\mathcal{I}_{E}
\end{cases}
\end{align*}
where $g_{t,k}:=\nabla F_{k}(w_{t}^{k},\xi_{t}^{k})$ is the stochastic
gradient. 
\begin{theorem}
	Assume that $F_{k}$ is $L$-smooth and $\mu$-strongly convex for
	all $k$, and let $\nu_{\max}=\max_{k}\frac{1}{N}p_{k}$. Let $\kappa=\frac{L}{\mu}$,
	$\gamma=\max\{32\kappa,E\}$ where $E$ is the communication delay,
	and diminishing learning rates $\alpha_{t}=\frac{9}{\mu}\frac{1}{t+\gamma}$
	and $\beta_{t-1}=\frac{9}{14\max\{\mu,1\}}\frac{1}{t+\gamma}$. Let
	$\overline{y}_{T}=\sum_{k=1}^{N}p_{k}y_{T}^{k}$ be the average of
	local parameters at an arbitrary time $T$.

@@

In this section, we provide linear speedup analyses of Nesterov Accelerated
FedAvg with convex and smooth objectives and show that the optimality
gap is $O(1/\sqrt{NT})$ where $N$ is the number of participating
devices. This result complements the strongly convex case in the previous
part, as well as the non-convex smooth setting in {[}CITE kdd and
others{]}, where a similar $O(1/\sqrt{NT})$ rate is given in terms
of averaged gradient norm. 
\begin{theorem}
	Suppose $F_{k}$ is $L$-smooth and convex for all $k$, and let $\nu_{\max}=\max_{k}\frac{1}{N}p_{k}$.
	Set learning rate $\alpha_{t},\beta_{t}=O(\sqrt{\frac{N}{T}})$.

	With full device participation, 
	\begin{align*}
	\min_{t\leq T}F(\overline{y}_{t})-F(w^{\ast}) & =O(\sqrt{\frac{\nu_{\max}^{2}\sigma^{2}}{NT}+\frac{E^{2}LG^{2}}{T^{4/3}}})
	\end{align*}
	and with partial device participation with $K$ sampled devices at
	each communication round, 
	\begin{align*}
	\min_{t\leq T}F(\overline{y}_{t})-F(w^{\ast}) & =O(\sqrt{\frac{\nu_{\max}^{2}\sigma^{2}}{NT}+\frac{E^{2}G^{2}}{KT}+\frac{E^{2}LG^{2}}{T^{4/3}}})
	\end{align*}
\end{theorem}
%
\begin{remark}
	It is possible to extend the analysis to accelerated FedAvg algorithms
	with other momentum-based updates. However, in line with the stochastic
	optimization setting, none of these methods will be able to achieve
	a better rate than the original FedAvg algorithm in general. For this
	reason, we will instead turn to the overparameterized setting {[}CITE
	Belkin{]} where we show that it is possible to improve the convergence
	rate of FedAvg with momentum based local updates. 
\end{remark}
